Integer Programming Formulations for the Shared Multicast Tree Problem

### 2018-09-24 smt/sec2-literature.tex
DH: Minor revision of most files.

\section{Literature Overview}
\label{sec:literature}
The combinatorial optimization literature is already rich on Integer Programming (IP) approaches to power minimization in wireless networks.
Since the work of \citet{Wieseltier00onthe}, in particular the MEB and MEM problems have been studied extensively.
Noteworthy contributions to this research include the IP models studied by \citet{das03},
\citet{altinkemer05}, \citet{yuan05}, \citet{yuan08}, \citet{bauer08}, \citet{montemanni11} and \citet{leggieri08}. 
\textcolor{blue}{
The latter paper proposes an IP formulation of MEM based on a Set Covering model together with a preprocessing procedure reducing the number of constraints.}
\citet{cagalj02} proved that MEB, and thereby MEM, is NP-hard.
Numerous contributions to fast, but possible inexact methods for MEM have thus appeared, and the interested reader is referred to \citep{hsiao13} for an overview. 
\textcolor{blue}{Another study concerning MEM is \cite{guo}, where the authors investigate energy conservation while using adaptive antennas. 
Unlike omnidirectional antennas, adaptive antennas can concentrate their transmission power in a smaller angle.}

The SBT problem has also received some attention in the scientific literature.
It was introduced by \citet{Papadimitriou06SBT}, who also proved that the problem is NP-hard.
In the same article, an approximation algorithm for SBT was developed and analyzed experimentally.
Building on this work, \citet{Haugland12Dual} presented an integer programming model for SBT, along with an associated decomposition approach.

>>>>

\section{Literature Overview}
\label{sec:literature}

The combinatorial optimization literature is already rich on Integer Programming (IP) approaches to power minimization in wireless networks.
Since the work of \citet{Wieseltier00onthe}, in particular the MEB and MEM problems have been studied extensively.
Noteworthy contributions to this research include the IP models studied by \citet{das03},
\citet{altinkemer05}, \citet{yuan05}, \citet{yuan08}, \citet{bauer08}, \citet{montemanni11} and \citet{leggieri08}. 
\textcolor{blue}{
The latter paper proposes an IP formulation of MEM based on a set covering model together with a preprocessing procedure reducing the number of constraints.}
\citet{cagalj02} prove that MEB, and thereby MEM, is NP-hard.
Numerous contributions to fast, but possible inexact methods for MEM have thus appeared, and the interested reader is referred to \citep{hsiao13} for an overview. 
\textcolor{blue}{Another study concerning MEM is the work by \cite{guo}, who investigate energy conservation while using adaptive antennas. 
Unlike omnidirectional antennas, adaptive antennas can concentrate their transmission power in a smaller angle.}

\textcolor{blue}{\citet{bein10} study energy efficiency in wireless networks where all nodes can initiate a message, and the communication takes place in a single all-to-all broadcast tree.
A heuristic algorithm generating all-to-all broadcast trees is proposed by \citet{bhukya14}, along with an experimental evaluation confirming its merit.}

The SBT problem has also received some attention in the scientific literature.
It was first introduced by \citet{Papadimitriou06SBT}, who also proved that the problem is NP-hard.
In the same article, an approximation algorithm for SBT is developed and analyzed experimentally.
Building on this work, \citet{Haugland12Dual} present an integer programming model for SBT, along with an associated decomposition approach.

### 2018-09-24 smt/sec3-sbtsmt.tex
DH: Minor revision of most files.

A WANET is modeled by a complete graph $G=(V,E)$, where the set $V$ of nodes represents the set of wireless devices and the set of edges $E=\{\{i,j\}:i,j\in V, i\neq j\}$ corresponds to the potential links between them.
The set $A=\{(i,j):i,j\in V,\{i,j\}\in E\}$ consists of all arcs that can be derived by directing the edges in $E$.
Nodes in the set $D\subseteq V$ are the destination nodes, also referred to as \emph{source} nodes.
For an arbitrary $i\in V$, sets $V\setminus \{i\}$ and $D\setminus\{i\}$ are denoted $V_i$ and $D_i$, respectively.

Sending a message directly from (to) node $i$ to (from) node $j$ requires a certain \emph{power} $p_{ij}=p_{ji}$.
Whenever appropriate, $p_{ij}$ is referred to as the \emph{cost} of $\{i,j\}$,
and edges with higher cost than others are said to be \emph{more expensive}.
When the nodes are embedded in the plane, $p_{ij}$ is often assumed \citep{Papadimitriou06SBT,Haugland12Dual} to be
proportional to the Euclidean distance between $i$ and $j$ raised to a power between 2 and 4.
For simplicity, we assume for all $i\in V$ that $p_{ij}\neq p_{ik}$ when $j\neq k$, and define
$W_{ij}=\left\{k\in V_i: p_{ik}\geq p_{ij}\right\}$
as the set of nodes requiring no less power at $i$ than $j$ does.
The notation $G'\subseteq G$ means that $G'=(V',E')$ is a subgraph of $G$, i.e., $E'\subseteq E$ consists of the edges in $E$ with both end nodes in $V'\subseteq V$.

@@

In more general terms, let $T$ be a multicast tree, and let $i\in V_T$ be a node with at least two neighbors in $T$.
For multicasting a message, the power to be assigned to node $i$ is either $p_{ii_1}$ or $p_{ii_2}$, where $i_1$ and $i_2$ are neighbors of $i$ such that
$p_{ii_1}>p_{ii_2}>p_{ij}$ for all other neighbors $j\in V_T\setminus\{i_1,i_2\}$.
If the path from the message source to $i$ intersects $i_1$, then power $p_{ii_2}$ is sufficient.
Otherwise, the power $p_{ii_1}$ is required.
At a leaf $i$ with neighbor $i_1$ in $T$, the power assignment is $p_{ii_1}$ if $i$ also is the message source, and zero otherwise.

All nodes in $D$ are assumed to be equally likely to initiate a message.
The frequency at which power levels $p_{ii_1}$ and $p_{ii_2}$ (let $p_{ii_2}=0$ if $i$ is a leaf) are required,
is thus proportional to the number of sources in $D$ that utilize the respective arc for message forwarding.
In the example in Fig.\ \ref{fig:objexp}, the expected power consumption at node $i$ is thus proportional to $4p_{ii_1} + 3p_{ii_2}$.
Summing over all nodes yields the total cost of the tree.
% The objective in the problem addressed in the current work is to minimize this cost.

\textcolor{blue}{
Different energy models can be found in the literature. 
\cite{halgamuge} present a comprehensive energy model for wireless sensor networks, 
where the sources of energy consumption consist of sensing, logging, processing, transimission, actuation, transient and cluster formation.
Our energy model focuses on the energy consumed by transmission, which is also tho most energy consuming aspect of wireless networks (51\% according to \cite{halgamuge}).
Some of the measures such as sensing and cluster formation are not relevant in our model.
Another distinction is that nodes in our model are all equivalent and the data is not aggregated in the nodes.
On top of that, all the related works menioned in Section \ref{sec:literature} assume transmission as the sole energy consumption source. 
For these reasons we disregard the potential energy consumption sources occuring in sensor networks.}

\textcolor{blue}{
However, it is easy to incorporate some of the disregarded measures by adding a constant for every used Steiner node.}

@@

To express the cost of $T$ more rigorously, let $T^s=(V_T,A^s)$ denote the directed tree (arborescence) obtained by directing all edges in $E_T$ such that they point away from the \emph{root} at $s$.
For each arc $(i,j)\in A$, we let $n_{ij}(T)$ denote the number of destinations $s$ such that $(i,j)$ is the most expensive arc leaving node $i$ in $T^s$.
That is,
\[
  n_{ij}(T) = \left|\left\{s\in D: p_{ij}=\max_{k\in V_i: (i,k)\in A^s}p_{ik}\right\}\right|,
\]

>>>>

A WANET is modeled by a complete graph $G=(V,E)$, where the set $V$ of nodes represents the set of wireless devices,
and the set of edges $E=\{\{i,j\}:i,j\in V, i\neq j\}$ corresponds to the potential links between them.
The set $A=\{(i,j):i,j\in V,\{i,j\}\in E\}$ consists of all arcs that can be derived by directing the edges in $E$.
Nodes in the set $D\subseteq V$ are the destination nodes, also referred to as \emph{source} nodes.
For an arbitrary $i\in V$, sets $V\setminus \{i\}$ and $D\setminus\{i\}$ are denoted $V_i$ and $D_i$, respectively.

Sending a message directly from (to) node $i$ to (from) node $j$ requires a certain \emph{transmission power} $p_{ij}=p_{ji}$.
Whenever appropriate, $p_{ij}$ is referred to as the \emph{cost} of $\{i,j\}$,
and edges with higher cost than others are said to be \emph{more expensive}.
When the nodes are embedded in the plane, $p_{ij}$ is often assumed \citep{Papadimitriou06SBT,Haugland12Dual,halgamuge} to be
proportional to the Euclidean distance between $i$ and $j$ raised to a power between 2 and 4.
For simplicity, we assume for all $i\in V$ that $p_{ij}\neq p_{ik}$ when $j\neq k$, and define
$W_{ij}=\left\{k\in V_i: p_{ik}\geq p_{ij}\right\}$
as the set of nodes requiring no less power at $i$ than $j$ does.
The notation $G'\subseteq G$ means that $G'=(V_{G'},E_{G'})$ is a subgraph of $G$, i.e., $E_{G'}\subseteq E$ consists of the edges in $E$ with both end nodes in $V_{G'}\subseteq V$.

@@

In more general terms, let $T\subseteq G$ be a multicast tree, and let $i\in V_T$ be a node with at least two neighbors in $T$.
For multicasting a message, the power to be assigned to node $i$ is either $p_{ii_1}$ or $p_{ii_2}$, where $i_1$ and $i_2$ are neighbors of $i$ such that
$p_{ii_1}>p_{ii_2}>p_{ij}$ for all other neighbors $j\in V_T\setminus\{i_1,i_2\}$.
If the path from the message source to $i$ intersects $i_1$, then power $p_{ii_2}$ is sufficient.
Otherwise, the power $p_{ii_1}$ is required.
At a leaf $i$ with neighbor $i_1$ in $T$, the power assignment is $p_{ii_1}$ if $i$ also is the message source, and zero otherwise.

All nodes in $D$ are assumed to be equally likely to initiate a message.
The frequency at which power levels $p_{ii_1}$ and $p_{ii_2}$ (let $p_{ii_2}=0$ if $i$ is a leaf) are required,
is thus proportional to the number of sources in $D$ that utilize the respective arc for message forwarding.
In the example in Fig.\ \ref{fig:objexp}, the expected power consumption at node $i$ is thus proportional to $4p_{ii_1} + 3p_{ii_2}$.
Summing over all nodes yields the total cost of the tree.

\textcolor{blue}{
The cost model outlined above aims for minimization of the transmission energy.
As pointed out by \citet{halgamuge}, transmission constitutes a substantial part of the total energy consumption in wireless sensor networks.
The energy consumed by other operations, such as sensing, logging, processing, actuation, and cluster formation is less sensitive to the topology
of the multicast tree, and is neglected in the current work.
This is in line with the approach taken in the wide range of literature on optimization models for wireless networks mentioned in Section \ref{sec:literature}.
For a detailed energy model for wireless sensor networks, the reader is referred to the article by \citet{halgamuge}.}

@@

To express the cost of the multicast tree more rigorously, let $T^s=(V_T,A^s)$ denote the directed tree (arborescence) obtained by directing all edges in $T$ such that they point away from the \emph{root} at $s$.
For each arc $(i,j)\in A$, we let $n_{ij}(T)$ denote the number of destinations $s$ such that $(i,j)$ is the most expensive arc leaving node $i$ in $T^s$.
That is,
\[
  n_{ij}(T) = \left|\left\{s\in D: p_{ij}=\max_{k\in V_i: (i,k)\in A^s}p_{ik}\right\}\right|,
\]

### 2018-09-24 smt/sec8-exp.tex
DH: Minor revision of most files.

\begin{algorithm}
\textcolor{blue}{
Create an empty pool $P$ of elite solutions with capacity $\gamma=\left\lceil\sqrt{M/2}\right\rceil$\;
\For {$i\in \{1,\dots,M$\}} {
Generate a new solution $S$ using a constructive algorithm, local search, and randomization\;
\If {$i<\gamma$} {$P\leftarrow P\cup \{S\}$}
\Else {
$\hat{S}\leftarrow\arg\max_{S'\in P} c(S')$\;
\If {$c(\hat{S})> c(S)$}{$P\leftarrow P\cup \{S\}\setminus\{\hat{S}\}$}
$\hat{S}\leftarrow\arg\max_{S'\in P} c(S')$\;
Generate a solution $\tilde{S}$ by combining $S$ with solutions in $P$\;
\If {$c(\hat{S})> c(\tilde{S})$}{$P\leftarrow P\cup \{\tilde{S}\}\setminus\{\hat{S}\}$}
}
}
\Return $\arg\min_{S'\in P} c(S')$\;
 \caption{Outline of the metaheuristic algorithm}
}
\label{alg:meta}
\end{algorithm}
\textcolor{blue}{The construction procedure and local search is the adapted BIP algorithm together with refinements presented in \citet{ivanova16isco}.
The remaining parts of the algorithm are carried out as discussed in \citet{pajor18}.
The randomization is achieved by perturbing the edge costs in the input graph.
The obtained solution $S$ is evaluated with respect to the real (original) edge costs and added to the pool $P$ if the pool capacity is not filled.
If the pool is full and its worst solution $\hat{S}$ is worse than $S$, $\hat{S}$ is replaced by $S$ in $P$.
Then a new solution $\tilde{S}$ is constructed by combining some solution in $P$ with $S$. 
First, we adjust edge costs resulting in a modified input instance, 
and then run the construction and local search yielding $\tilde{S}$.
This new solution is again added to the pool if it is better than some solution already in $P$.
The adjustment of the edges tries to promote edges that appear in both $S$ and a selected solution in $P$. 
In that case, the original cost is preserved.
If an edge is present in only one of the two solutions, its cost is multiplied by a random integer between 100 and 500.
If an edge appears in neither of the solutions, its edge is multiplied by 1000, which makes it very unlikely to appear in the result $\tilde{S}$ of the combination.
}
\begin{table}[]
\centering
\setlength{\tabcolsep}{6pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.4} % Default value: 1
\textcolor{blue}{
\begin{tabular}{rrrrrrrrrrr}
\multicolumn{1}{l}{}      & \multicolumn{1}{l}{}      & \multicolumn{3}{c}{5 min}  & \multicolumn{3}{c}{10 min} & \multicolumn{3}{c}{20 min}     \\\hline
\multicolumn{1}{l}{$|V|$} & \multicolumn{1}{l}{$|D|$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} \\
\hline
18  & 12   &100.0&100.0&101.8&100.0&100.0&100.0&100.0&100.0&100.0\\
21  & 14   &106.5&107.4&101.2&102.7&103.1&100.1&100.9&100.3&100.0\\
24  & 16   &144.2&112.2&101.3&124.0&114.0&100.1&104.3&112.7&100.0\\     
\hline
20  & 10   &100.6&100.0&101.4&100.1&100.0&100.0&100.1&100.0&100.0\\
22  & 11   &102.1&103.2&101.7&101.3&100.7&100.1&100.7&100.2&100.1\\
24  & 12   &106.6&111.7&101.7&105.0&105.5&100.0&103.1&106.2&100.0\\                               
\hline
21  & 7    &100.1&100.0&100.0&100.0&100.0&100.0&100.0&100.0&100.0\\
24  & 8    &102.3&101.4&102.0&102.1&100.7&100.1&101.1&100.1&100.0\\
27  & 9    &127.4&114.0&101.6&111.2&109.0&100.1&108.2&104.1&100.0                                  
\end{tabular}
}
\caption{\textcolor{blue}{Comparison of the metaheuristic algorithm and solving IP models with imposed time-limit.}}
\label{tab:metamod}
\end{table}

>>>>

\textcolor{blue}{The construction procedure and local search is the adapted BIP algorithm together with refinements presented in \citet{ivanova16isco}.
The remaining parts of the algorithm are carried out as discussed in \citet{pajor18}.
The randomization is achieved by perturbing the edge costs in the input graph.
The obtained solution $S$ is evaluated with respect to the real (original) edge costs and added to the pool $P$ if the pool capacity is not filled.
If the pool is full and its worst solution $\hat{S}$ is worse than $S$, $\hat{S}$ is replaced by $S$ in $P$.
Then a new solution $\tilde{S}$ is constructed by combining some solution in $P$ with $S$. 
First, we adjust edge costs resulting in a modified input instance, 
and then run the construction and local search yielding $\tilde{S}$.
This new solution is again added to the pool if it is better than some solution already in $P$.
The adjustment of the edges tries to promote edges that appear in both $S$ and a selected solution in $P$. 
In that case, the original cost is preserved.
If an edge is present in only one of the two solutions, its cost is multiplied by a random integer between 100 and 500.
If an edge appears in neither of the solutions, its edge is multiplied by 1000, which makes it very unlikely to appear in the result $\tilde{S}$ of the combination.
}
\begin{table}[]
\centering
\setlength{\tabcolsep}{6pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.4} % Default value: 1
\textcolor{blue}{
\begin{tabular}{rrrrrrrrrrr}
\multicolumn{1}{l}{}      & \multicolumn{1}{l}{}      & \multicolumn{3}{c}{5 min}  & \multicolumn{3}{c}{10 min} & \multicolumn{3}{c}{20 min}     \\\hline
\multicolumn{1}{l}{$|V|$} & \multicolumn{1}{l}{$|D|$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} & \multicolumn{1}{c}{$\mathcal{X}_1$} & \multicolumn{1}{c}{$\mathcal{F}_1$} & \multicolumn{1}{c}{$\mathcal{M}$} \\
\hline
18  & 12   &100.0&100.0&101.8&100.0&100.0&100.0&100.0&100.0&100.0\\
21  & 14   &106.5&107.4&101.2&102.7&103.1&100.1&100.9&100.3&100.0\\
24  & 16   &144.2&112.2&101.3&124.0&114.0&100.1&104.3&112.7&100.0\\     
\hline
20  & 10   &100.6&100.0&101.4&100.1&100.0&100.0&100.1&100.0&100.0\\
22  & 11   &102.1&103.2&101.7&101.3&100.7&100.1&100.7&100.2&100.1\\
24  & 12   &106.6&111.7&101.7&105.0&105.5&100.0&103.1&106.2&100.0\\                               
\hline
21  & 7    &100.1&100.0&100.0&100.0&100.0&100.0&100.0&100.0&100.0\\
24  & 8    &102.3&101.4&102.0&102.1&100.7&100.1&101.1&100.1&100.0\\
27  & 9    &127.4&114.0&101.6&111.2&109.0&100.1&108.2&104.1&100.0                                  
\end{tabular}
}
\caption{\textcolor{blue}{Comparison of the metaheuristic algorithm and solving IP models with imposed time-limit.}}
\label{tab:metamod}
\end{table}
